A Study on Common Bug Fix Patterns in Rust

### 2022-10-11 sections/methodology.tex
merge successful

The general idea for automatically finding the pervasive code patterns within a list of programs is to parse the programs and encode the important parts of the derived AST in fixed-size datapoints. Work in this vein makes the assumption that each cluster corresponds to one change pattern.

@@

To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs, encoded in two Python dictionaries. Tree diff algorithms can compute the differences between two arbitrary trees. If the trees are programs' abstract syntax trees, we call their diff an ASTDiff. We use dictdiff to obtain ASTDiffs.

@@

Our target repositories were the top 18 most-starred open source Rust projects on GitHub. We mined (using Pydriller) all the bug related commits and ran them through our pipeline. Then, we applied the DBSCAN clustering algorithm on the resulting datapoints. In Section~\ref{sec:clustering_data} we justify our choice of DBSCAN for clustering, and how we tuned it to improve cluster quality.

@@

As mentioned above, we use Syn for parsing Rust. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related lexemes (BCL) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCL+ elements can be found within Appendix~\ref{sec:appendix}. 
% I think we don't need this sentence here. Let's see what we do later.
%We define dimension items as \verb+DI = NT+ $\cup$ \verb+BCL+, as this set also equals the columns of our tables.

We parse two versions of a changed Rust file in a commit: the file before the commit and after it. This yields two Syn trees. Using the PLY tool (a Python implementation of lex and yacc), we wrote a simple transformer from the serialized Rust AST to Python dictionaries. The change that transforms the first tree to the second one is the fix that was applied in the commit. To find this difference, we use dictdiff, a Python library to find the diff of two Python dictionaries. Its output, in our context, is the ASTDiff.

@@

The output of dictdiff is a list of diffs, where each diff comprises three parts. The first part specifies the modification type:

@@

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one contributing element to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCL}$).

@@

\paragraph{Encoding the data.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. We need fixed-size datapoints, as they are a requirement for both centroid-based and density-based well-known clustering algorithms. The number of columns would depend on the amount of information we want to encode in the datapoints.

The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an encoding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to encode all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCL}$ as our dimensions, which yields an infeasible $O(n^n)$ dimensions; this could potentially be reduced somewhat, but is still impractical. While we can implement a specific workaround to distinguish nested ifs (and we have done so), the general point about the large dimensionality of fully order-sensitive encodings still holds.

A reasonable work around would be to reduce $n$ from the full set of non terminals to a smaller set of categories (e.g. a category for larger entites like class or function definitions, and a category for smaller entities like statements and expressions), and then to record all possible combinations of these categories (similar to JS authors~\cite{foo}). That approach reduces the number of dimensions and provides more reasonable and syntactically-correct combinations. However, it would still yield a sparse dataset. 
% todo you need to dig up the JS paper ref / Not sure, how to cite, I usually create a bib file, but i'm not sure if i'm allowed to do that for this template.

In our methodology, we propose a novel representation. Similar to the first approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCL}$ we collected from Syn. However, to simulate order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

We made two observations about changes. First, we saw that an item (non terminal or BC-related lexeme) that occurs closer to AST leaves tends to be more semantically important. Relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

@@

Furthermore, as we wanted to make sure that our tool captures patterns related to the Rust borrow checker, we manually increased the weights of the \texttt{BCL} items. The manual adjustment is why we characterize our weighting scheme as semi-automatically designed.

@@

We used the Pydriller library to mine software repositories. Pydriller provides APIs to extract commits from a Git repository and to search through different revisions of files. Algorithm~\ref{alg} shows how we used Pydriller to run the target repositories through our code analysis pipeline and populate two databases: one for borrow-checker related patterns (which involve items in \texttt{BCL}) and one for general patterns (which don't). For each repository, we search all commits that include bug-fixing related keywords within their commit messages. Here, we chose bug-fixing related keywords based on the set introduced in \cite{zhang2018empirical}, excluding `nan' and `inf'.

@@

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{figs/clusters.pdf}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters $\epsilon$ and $Z$, respectively separated in columns and rows. In each subplot, the two blue bars in the left indicate the number of clustered datapoints (the number is shown above each bar) after applying clustering on $D_g$ ($G_c$) and $D_b$ ($BC_{c}$), respectively. The two orange bars in the right show similar data but for noise datapoints. In each subplot's legend, $C_G$ and $C_B$ show the number of clusters obtained from $D_g$ and $D_b$, respectively}
\end{figure}

>>>>

The general idea for automatically finding the pervasive code patterns within a list of programs is to parse the programs and encode the important parts of the derived AST in fixed-size datapoints. Work in this vein makes the assumption that each cluster corresponds to one change pattern \cite{hanam2016discovering} \cite{campos2019discovering} \cite{yang2022mining}.

@@

To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs, encoded in two Python dictionaries. Tree diff algorithms can compute the differences between two arbitrary trees. If the trees are programs' abstract syntax trees, we call their diff an ASTDiff. We use dictdiffer to obtain ASTDiffs.

@@

Our target repositories were the top 18 most-starred open source Rust projects on GitHub. We mined (using Pydriller \cite{spadini2018pydriller}) all the bug related commits and ran them through our pipeline. Then, we applied the DBSCAN clustering algorithm \cite{ester1996density} on the resulting datapoints. In Section~\ref{sec:clustering_data} we justify our choice of DBSCAN for clustering, and how we tuned it to improve cluster quality.

@@

As mentioned above, we use Syn for parsing Rust. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (BCE) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCE+ elements can be found within Appendix~\ref{sec:appendix}.

We parse two versions of a changed Rust file in a commit: the file before the commit and after it. This yields two Syn trees. Using the PLY tool\footnote{\url{https://www.dabeaz.com/ply/}} (a Python implementation of lex and yacc), we wrote a simple transformer from the serialized Rust AST to Python dictionaries. The change that transforms the first tree to the second one is the fix that was applied in the commit. To find this difference, we use dictdiffer\footnote{\url{https://dictdiffer.readthedocs.io/en/latest/}}, a Python library to find the diff of two Python dictionaries. Its output, in our context, is the ASTDiff.

@@

The output of dictdiffer is a list of diffs, where each diff comprises three parts. The first part specifies the modification type:

@@

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one contributing element to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCE}$).

@@

\paragraph{Encoding the data.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. We need fixed-size datapoints, as they are a requirement for both centroid-based and density-based well-known clustering algorithms \cite{xu2005survey}. The number of columns would depend on the amount of information we want to encode in the datapoints.

The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an encoding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to encode all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCE}$ as our dimensions, which yields an infeasible $O(n^d)$ dimensions ($n=|\mathtt{NT} \cup \mathtt{BCE}|$ and $d$ is the depth of nested structures); this could potentially be reduced somewhat, but is still impractical. While we can implement a specific workaround to distinguish nested ifs (and we have done so), the general point about the large dimensionality of fully order-sensitive encodings still holds.

A reasonable work around would be to reduce $n$ from the full set of non terminals to a smaller set of categories (e.g. a category for larger entites like class or function definitions, and a category for smaller entities like statements and expressions), and then to record all possible combinations of these categories (similar to \cite{hanam2016discovering}). That approach reduces the number of dimensions and provides more reasonable and syntactically-correct combinations. However, it would still yield a sparse dataset.

In our methodology, we propose a novel representation. Similar to the first approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn. However, to simulate order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

We made two observations about changes. First, we saw that an item (non terminal or BC-related element) that occurs closer to AST leaves tends to be more semantically important. Relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

@@

Furthermore, as we wanted to make sure that our tool captures patterns related to the Rust borrow checker, we manually increased the weights of the \texttt{BCE} items. The manual adjustment is why we characterize our weighting scheme as semi-automatically designed.

@@

We used the Pydriller library to mine software repositories. Pydriller provides APIs to extract commits from a Git repository and to search through different revisions of files. Algorithm~\ref{alg} shows how we used Pydriller to run the target repositories through our code analysis pipeline and populate two databases: one for borrow-checker related patterns (which involve items in \texttt{BCE}) and one for general patterns (which don't). For each repository, we search all commits that include bug-fixing related keywords within their commit messages. Here, we chose bug-fixing related keywords based on the set introduced in \cite{zhang2018empirical}, excluding `nan' and `inf'.

@@

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{figs/clusters.pdf}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters $\epsilon$ and $Z$, respectively separated in columns and rows. In each subplot, the two blue bars in the left respectively indicate the number of clustered and noise datapoints (the number is shown above each bar) after applying clustering on $D_g$ (denoted with $G_c$ and $G_n$ on the x axis). The two orange bars in the right show similar data but for the $D_{b}$ database. In each subplot's legend, $C_G$ and $C_B$ show the number of clusters obtained from $D_g$ and $D_b$, respectively}
\end{figure}
